---
title: "Class09 PCA and CLustering Project"
author: "Lydia Hernandez"
date: "2/8/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read our input data

```{r}
fna.data <- "WisconsinCancer.csv"


# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data)
```
Always double check your data
```{r}
head(wisc.df)
```

Looks like there is a funny last column "X". Lets check how many samples (i.e. patients) and features (i.e. cols) are in this data.

```{r}
# number of patients
nrow(wisc.df)
```

```{r}
# number of features
ncol(wisc.df)
```

Lets take cols 3 to 32 (i.e. drop the funny "X" and the diagnosis cols)
```{r}
wisc.data <- wisc.df[ ,3:32]
head(wisc.data)
```

Add patientID as row names of our data

```{r}
rownames(wisc.data) <- wisc.df$id
head(wisc.data)
```

How many cancer/non-cancer samples?
```{r}
table(wisc.df$diagnosis)
```

How many features are "_mean" values?

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)))
```

```{r}
inds <- grep("_mean", colnames(wisc.data))
colnames(wisc.data)[inds]
```

Check mean and standard deviation of features
```{r}
# Check column means and standard deviations

colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
# apply the mean of every column column is the number 2 here the second 2 is sig figs
round(apply(wisc.data,2, mean), 2)
```

```{r}
round(apply(wisc.data,2,sd), 2)
```

Looks like we need to scale data

## Principal Component Analysis (PCA)

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

```{r}
biplot(wisc.pr)
```

Nope! That is a useless plot

We need to make our own plot of our PCA results
```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1], wisc.pr$x[,2])
```

Lets use the expert pathologist diagnosis to color our plot

```{r}
wisc.df$diagnosis
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = wisc.df$diagnosis, xlab = "PC1", ylab = "PC2")
```

### Make a scree plot for our PCA results

This plot will show the proportion of variance captured in each PC.

```{r}
variance <- wisc.pr$sdev^2
pve <- round(variance/sum(variance) *100, 2)
```

```{r}
plot(pve, type = "o")
```

```{r}
barplot(pve, axes = FALSE, names.arg = paste("PC", 1:length(pve), sep=""), las =2, ylab = "Proportion of Variance (%)")
axis(2, round(pve))
```

## Clustering in Principal Component Space

For hclus we need a distance matri and we gt this from our PCA results (i.e. wisc.pr$x)

```{r}
d <- dist(wisc.pr$x[, 1:2])
hc <- hclust(d, method = "ward.D2")
plot(hc)
```

Let's cut or tree into 3 groups (i.e. clusters!)

```{r}
grp3 <- cutree(hc, k=3)
table(grp3)
```

Plot our PCA plot colored by cluster

```{r}
plot(wisc.pr$x[, 1], wisc.pr$x[,2], xlab = "PC1", ylab = "PC2",
     col = grp3)
```


Cross Tabulation

```{r}
#table(grp3, )
#diagnosis

diagnosis <- wisc.df$diagnosis == "M"
table(grp3, wisc.df$diagnosis)
```

```{r}
new <- read.csv("new_samples.csv")
npc <- predict(wisc.pr, newdata=new)
npc
```


```{r}
plot(wisc.pr$x[,1:2], col=wisc.df$diagnosis)
points(npc[,1], npc[,2], col=c("green", "blue"), pch=15, cex=2)
```








