---
title: "Class 14: Transcriptomics and the analysis of RNA-Seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```

How many genes are in this dataset?
```{r}
nrow(counts)
```

How many exps?
```{r}
ncol(counts)
```


Let's check for metadata to counts correspondence
```{r}
#all command summarizes results "true" equals "all true"
all(metadata$id == colnames(counts)[-1])
```

Extracting control exps data (i.e. columns of our **counts** object based on the **metadata**)

```{r}
#find the column names (ids) of the 'control' experiments
control.ids <- metadata$id[metadata[,"dex"] == "control"]
control.ids

#use the ids to extract the control counts columns
control <- counts[,control.ids]
rownames(control) <- counts$ensgene
head(control)
```


```{r}
#find the column names (ids) of the 'treated' experiments
treated.inds <- metadata$id[metadata[,"dex"] == "treated"]
treated.inds

#use the ids to extract the treated counts columns
treated <- counts[,treated.inds]
rownames(treated) <- counts$ensgene
head(treated)
```

Find the mean counts value for each gene in  control experiments
```{r}
control.mean <- rowSums(control)/ncol(control)
head(control.mean)
```

Find the mean counts value for each gene in treated experiments
```{r}
treated.mean <- rowSums(treated)/ncol(treated)
head(treated.mean)
```

Combine these for ease of use later
```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
colSums(meancounts)
```

Let's plot it! 
```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

make a log-log plot

```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log = "xy")
```

We can find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples. We usually look at the log2 of the fold change, because this has better mathematical properties.

Here we calculate log2foldchange, add it to our meancounts data.frame and inspect the results either with the head() or the View() function for example.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

We need to filter out the zero value entry genes
```{r}
# meancounts[,1:2]

toy <- c(1,2,0,4,0)
toy == 0
which(toy == 0)

newtoy <- cbind(toy, toy)
newtoy 

which(newtoy == 0, arr.ind=TRUE)
# which rows contain zero values?
unique(which(newtoy == 0, arr.ind=TRUE)[,1])
```


```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
head(zero.vals)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
```

```{r}
sum(down.ind)
```

```{r}
head(mycounts[up.ind,])
```

## Add annotation data from Ensembl annotation file

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

```{r}
#how much stuff is in this dataset? answer given in number of rows and columns 
dim(anno)
```

Let's try out the **merge()** function to add the annotation data to our **mycounts** results
```{r}
head(mycounts)
```

```{r}
mycounts.anno <- merge(x=mycounts, y=anno, by.x = "row.names", by.y = "ensgene")
head(mycounts.anno)
```
```{r}
library("AnnotationDbi")
```

```{r, eval = FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("AnnotationDbi")
biocLite("org.Hs.eg.db")
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```


```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
head(mycounts)
```

# DESeq2 analysis

```{r}
library(DESeq2)
```

Build our DESeqDataSet required for DESeq analysis
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

Run the DESeq analysis
```{r}
dds <- DESeq(dds)
```

Get our DESeq results
```{r}
res <- results(dds)
res
```

```{r}
summary(res, alpha=0.01)
```



```{r}
res$sig <- res$padj<0.05 & abs(res$log2FoldChange)>2

plot( res$log2FoldChange,  -log(res$padj), 
 col=res$sig+1, ylab="-Log(P-value)", xlab="Log2(FoldChange)")
# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.1), col="darkgray", lty=2)
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

#Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

